Review Paper Draft

### 2020-07-10 main.tex
A few notes from 2020-07-09 meeting about including unpenalized covariates in the model.

To describe and compare PC-Lasso and LMM-Lasso, we consider an $n \times p$ matrix of genotype data $\boldsymbol{G}$ for $n$ subjects and $p$ SNPs, where each $G_{i,j} \in \{ 0, 1, 2 \}$ enumerates the number of allele copies subject $i$ has for SNP $j$. We denote the normalized version of this genotype matrix as $\bX$, where each $X_{ij} = (G_{ij} - 2 p_j) / \sqrt{p_j (1 - p_j)}$ and $p_j$ is the minor allele frequency (MAF) of SNP $j$ \citep{zhang2015principal, price2006principal}. We assume that an $n \times 1$ vector of continuous phenotype values, $\by$, can be expressed as the sum of additive SNP effects, environmental confounding, and observational noise and consider the linear mixed model,
\begin{equation}
    \label{eqn:model}
    \by = \Int + \bX\bbeta + \bZ\bgamma + \beps
\end{equation}
where $\Int$ is the intercept, $\bbeta$ is a $p \times 1$ vector of SNP effects, $\bgamma$ is a $q \times 1$ vector of environmental effects, $\bZ$ is an $n \times q$ matrix that allocates environmental effects, and $\beps \sim N(\boldsymbol{0}, \sigmaee \bI_n)$. Often it is assumed that subjects belong to $q \le p$ subpopulations, and $\bZ$ is a corresponding incidence matrix of 0s and 1s. However, as we will discuss, this is not the only possibility.

@@

%------------------------------%
\subsection{LMM-Lasso}
LMM-Lasso models $\bZ\bgamma$ as a random effect assuming 
$$\bgamma \sim N\left(\boldsymbol{0}, \frac{\sigmagg}{p} \bI_p \right).$$  Typically, $\bZ$ is not observed directly, and a common convention is to assume $\bZ = \bX$ \citep{wang2018multiplex, lippert2011fast, yang2014advantages}.This results in the LMM regression model
\begin{equation}
    \label{eqn:lmm_reg}
    \by = \Int + \bX\bbeta + \bX\bgamma+ \beps,
\end{equation}
where we assume $\beps \independent \bX\bgamma$. Rather than assuming anything about $\bgamma$ directly, an equivalent representation lets $\bu=\bX\bgamma$ such that $\bu \sim N(\boldsymbol{0}, \sigmagg \bK)$, and yields the equivalent model
\begin{equation}
    \label{eqn:lmm_reg_u}
    \by = \Int + \bX\bbeta + \bu + \beps.
\end{equation}
Regardless of representation, the LMM assumes $\by \sim N(\Int + \bX \bbeta, \bV)$, where $\bV = \sigmagg \textbf{K} + \sigmaee \textbf{I}$. LMM-Lasso performs a weighted rotation of the original data, left-multiplying by $\bV^{-1/2}$ in a manner akin to generalized least squares to diagonalize the error term such that
\begin{equation}
\bV^{-\frac{1}{2}}\by \sim N(\bV^{-\frac{1}{2}} \left(\Int +\bX \bbeta\right), \bI).
\end{equation}
\anna{...however, as Karl Rohe paper shows, GLS does not completely translate to penalized regression...comment on this?}
\begin{comment}
For ease of notation, we incorporate the intercept into the design matrix and let $\bX' = \begin{bmatrix} \textbf{1} & \bX \end{bmatrix}$ and $\bbeta' = \begin{pmatrix} \mu & \bbeta \end{pmatrix}^T$ so that we can express this more compactly as 
\begin{equation}
\bV^{-\frac{1}{2}}\by \sim N(\bV^{-\frac{1}{2}} \bX' \bbeta', \bI).
\end{equation}
\end{comment}
Using the spectral decomposition of $\bK$ described in equation \eqref{eqn:k}, we can gain further insight into the mechanism of this transformation. Note that $\bV^{-1}$ can be expressed using the following matrix factorization 
\begin{align*}
    \bV^{-1} &= (\sigmagg \bK + \sigmaee \bI)^{-1}\\
    &=(\sigmagg \bU \bD^2 \bU^T + \sigmaee \bU \bU^T)^{-1}\\
    &= (\bU (\sigmagg \bD^2 + \sigmaee \bI) \bU^T)^{-1}\\
    &=\bU (\sigmagg \bD^2 + \sigmaee \bI)^{-1} \bU^T.
\end{align*}
By incorporating these decompositions, the transformed data can be expressed as $\bW \bXT$ and $\bW \byT$, respectively, where $\bXT = \bU^T \bX$ and $\byT = \bU^T \left(\by - \Int \right)$, where we center the transformed $\by$ to simplify notation and computation.  $\bW = \text{diag}\{w_1, ... w_n\}$ where $w_i$ corresponds to the square root of the $i$th element of the diagonal matrix $(\sigmagg \bD^2 + \sigmaee \bI)^{-1}$. Together, this yields the objective function
\begin{equation}
\label{eqn:lmm_obj}
Q(\bbeta | \bXT, \byT, \bW) = || \bW (\byT - \bXT \bbeta)||_2^2 + \lambda || \bbeta ||_1.
\end{equation}
An advantage to modeling unobserved confounding using a random effect, as opposed to fixed effects as in PC-Lasso, is that this requires the estimation of only $p + 3$ parameters.

@@

>>>>

\anna{We show this in the absence of fixed covariates for the sake of a clear illustration.. $\bX_0 \bbeta_0$ unpenalized things}

To describe and compare PC-Lasso and LMM-Lasso, we consider an $n \times p$ matrix of genotype data $\boldsymbol{G}$ for $n$ subjects and $p$ SNPs, where each $G_{i,j} \in \{ 0, 1, 2 \}$ enumerates the number of allele copies subject $i$ has for SNP $j$. We denote the normalized version of this genotype matrix as $\bX$, where each $X_{ij} = (G_{ij} - 2 p_j) / \sqrt{p_j (1 - p_j)}$ and $p_j$ is the minor allele frequency (MAF) of SNP $j$ \citep{zhang2015principal, price2006principal}. We assume that an $n \times 1$ vector of continuous phenotype values, $\by$, can be expressed as the sum of additive SNP effects, environmental confounding, and observational noise and consider the linear mixed model,
\begin{equation}
    \label{eqn:model}
    \by = \text{\anna{unpenalized stuff}} + \bX\bbeta + \bZ\bgamma + \beps
\end{equation}
where $\Int$ is the intercept, $\bbeta$ is a $p \times 1$ vector of SNP effects, $\bgamma$ is a $q \times 1$ vector of environmental effects, $\bZ$ is an $n \times q$ matrix that allocates environmental effects, and $\beps \sim N(\boldsymbol{0}, \sigmaee \bI_n)$. Often it is assumed that subjects belong to $q \le p$ subpopulations, and $\bZ$ is a corresponding incidence matrix of 0s and 1s. However, as we will discuss, this is not the only possibility.

@@

%------------------------------%
\subsection{LMM-Lasso}

\anna{We show this in the absence of fixed covariates for the sake of a clear illustration.. $\bX_0 \bbeta_0$ unpenalized things}

LMM-Lasso models $\bZ\bgamma$ as a random effect assuming 
$$\bgamma \sim N\left(\boldsymbol{0}, \frac{\sigmagg}{p} \bI_p \right).$$  Typically, $\bZ$ is not observed directly, and a common convention is to assume $\bZ = \bX$ \citep{wang2018multiplex, lippert2011fast, yang2014advantages}.This results in the LMM regression model
\begin{equation}
    \label{eqn:lmm_reg}
    \by = \Int + \bX\bbeta + \bX\bgamma+ \beps,
\end{equation}
where we assume $\beps \independent \bX\bgamma$. Rather than assuming anything about $\bgamma$ directly, an equivalent representation lets $\bu=\bX\bgamma$ such that $\bu \sim N(\boldsymbol{0}, \sigmagg \bK)$, and yields the equivalent model
\begin{equation}
    \label{eqn:lmm_reg_u}
    \by = \Int + \bX\bbeta + \bu + \beps.
\end{equation}
Regardless of representation, the LMM assumes $\by \sim N(\Int + \bX \bbeta, \bV)$, where $\bV = \sigmagg \textbf{K} + \sigmaee \textbf{I}$. LMM-Lasso performs a weighted rotation of the original data, left-multiplying by $\bV^{-1/2}$ in a manner akin to generalized least squares to diagonalize the error term such that
\begin{equation}
\bV^{-\frac{1}{2}}\by \sim N(\bV^{-\frac{1}{2}} \left(\Int +\bX \bbeta\right), \bI).
\end{equation}
\anna{...however, as Karl Rohe paper shows, GLS does not completely translate to penalized regression...comment on this?}
\begin{comment}
For ease of notation, we incorporate the intercept into the design matrix and let $\bX' = \begin{bmatrix} \textbf{1} & \bX \end{bmatrix}$ and $\bbeta' = \begin{pmatrix} \mu & \bbeta \end{pmatrix}^T$ so that we can express this more compactly as 
\begin{equation}
\bV^{-\frac{1}{2}}\by \sim N(\bV^{-\frac{1}{2}} \bX' \bbeta', \bI).
\end{equation}
\end{comment}
Using the spectral decomposition of $\bK$ described in equation \eqref{eqn:k}, we can gain further insight into the mechanism of this transformation. Note that $\bV^{-1}$ can be expressed using the following matrix factorization 
\begin{align*}
    \bV^{-1} &= (\sigmagg \bK + \sigmaee \bI)^{-1}\\
    &=(\sigmagg \bU \bD^2 \bU^T + \sigmaee \bU \bU^T)^{-1}\\
    &= (\bU (\sigmagg \bD^2 + \sigmaee \bI) \bU^T)^{-1}\\
    &=\bU (\sigmagg \bD^2 + \sigmaee \bI)^{-1} \bU^T.
\end{align*}
By incorporating these decompositions, the transformed data can be expressed as $\bW \bXT$ and $\bW \byT$, respectively, where $\bXT = \bU^T \bX$ and $\byT = \bU^T \left(\by - \Int \right)$, where we center the transformed $\by$ to simplify notation and computation.  $\bW = \text{diag}\{w_1, ... w_n\}$ where $w_i$ corresponds to the square root of the $i$th element of the diagonal matrix $(\sigmagg \bD^2 + \sigmaee \bI)^{-1}$. Together, this yields the objective function
\begin{equation}
\label{eqn:lmm_obj}
Q(\bbeta | \bXT, \byT, \bW) = || \bW (\byT - \bXT \bbeta)||_2^2 + \lambda || \bbeta ||_1.
\end{equation}
An advantage to modeling unobserved confounding using a random effect, as opposed to fixed effects as in PC-Lasso, is that this requires the estimation of only $p + 3$ parameters.

\anna{In the presence of unpenalized covariates to adjust for like age and sex... equations 7 and 8 and X-tilde become this... penalty term only penalizes beta1 - betas corresponding to SNPs}

@@

\anna{put derivation comparing lmmlasso and pclasso back}
